---
title: "Untitled"
output: html_document
date: "2023-12-24"
---
Problem 1: Probability, Distributed
1. Plot this function over the range −10 < x < +10, using an appropriate plot type.
```{r }
x=seq(-10, 10, 0.01)

pdf=rep(0, length(x))
pdf[x>-10 & x<10]=0.5 *exp(-abs(x)[x>-10 & x<10])  
plot(x, pdf, type='l')
```
```{r cars}
x= seq(-10, 10, 0.01)
cdf1= rep(0, length(x))
cdf1[x>0]= 1-0.5*exp(-x[x>0])       
cdf1[x <=0]=0.5*exp(x[x <=0])
plot(x, cdf1, type='l')
```
Problem 2: Pollution and Health
Q1.

```{r }
poll_df=read.csv('pollution.csv')
head(poll_df)

#unique(poll_df$area)
```
```{r }
df_control= poll_df[poll_df$area == 'C',]
df_indu= poll_df[poll_df$area== 'I', ]
par(mfrow=c(1,2))
hist(df_control$concentration, xlim=c(0, 15), ylim = c(0, 15), breaks=seq(0, 16, 1) , main= 'Area Control')
hist(df_indu$concentration, xlim=c(0, 15), ylim = c(0, 15), breaks=seq(0, 16, 1) , main= 'Area Industry')

```
Q-2

```{r }
pseudo.C.avg = unname(tapply(df_control$concentration, df_control$id, mean))
t.test(pseudo.C.avg)

pseudo.I.avg = unname(tapply(df_indu$concentratio, df_indu$id, mean))
t.test(pseudo.I.avg)
```

```{r }
mean_function = function(data, indices) {
  d = data[indices]
  return(mean(d))
}
library(boot)
boot_control = boot(df_control$concentration, mean_function, R = 1000)
boot_ci = boot.ci(boot_control, type = "basic") 
boot_ci
```
```{r }
boot_indu = boot(df_indu$concentration, mean_function, R = 1000)
boot_ci = boot.ci(boot_indu, type = "basic") 
boot_ci
```
Q-3
```{r }
var_test= var.test(df_indu$concentration, df_control$concentration)
var_test$p.value
var_test
```
Q-4

```{r }
#pairwise_test <- pairwise.t.test(poll_df$concentration, poll_df$area, p.adj = "bonferroni")
pairwise_test
```
```{r }
#t_test = t.test(df_control$concentration, df_indu$concentration)
t_test
```
```{r }
t.test(pseudo.I.avg ,pseudo.C.avg )

```
Q-5
```{r }
num_above_15 = sum(df_indu$concentration >= 15)
total_num = length(df_indu$concentration)
prop = num_above_15/total_num
prop
```
Q-6

```{r }
#calculate variance for group
var_cont = (sd(df_control$concentration)/0.3)^2
var_indu = (sd(df_indu$concentration)/0.3)^2
var_cont
var_indu
```

```{r }
df_os=read.csv('oscillate.csv')
head(df_os)
```
```{r}
lm_model_0 = lm(D ~ 1, data = df_os)  
lm_model_1 = lm(D ~ sin(w) + cos(w), data = df_os)
lm_model_2 = lm(D ~ sin(w) + cos(w) + sin(2*w) + cos(2*w), data = df_os)
lm_model_3 = lm(D ~ sin(w) + cos(w) + sin(2*w) + cos(2*w) + sin(3*w) + cos(3*w), data = df_os)
r1 = anova(lm_model_0, lm_model_1)
#summary (r1)
```
```{r }
r2= anova(lm_model_1, lm_model_2)
summary(r2)

```       
```{r }
r3= anova(lm_model_2, lm_model_3)
summary(r3)


```
```{r }
```

```{r }
library(ggplot2)
ggplot(poll_df, aes(x = concentration, fill = area)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 20) +
  labs(x = "Concentration", y = "Frequency", title = "Histogram of Concentrations by Area")

```
